NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-04-19 main.tex

NMRlipids Databank is a community-driven catalogue containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. It has been designed to improve the Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} of MD simulation data, most importantly the output trajectories. The NMRlipids Databank is constructed using the NMRlipids project protocol, in which all the content is openly accessible throughout the project~\cite{botan15}. Currently, the NMRlipids Databank contains 726 simulation trajectories with the total length of approximately 0.4\,ms. Single-component lipid membranes and binary mixtures, depicted in Fig.~\ref{fig:overlay}F, are currently most abundant in the NMRlipids Databank, yet mixtures with up to five lipid types are available. The distribution of lipids among the available simulations, shown in Fig.~\ref{fig:overlay}B, roughly resembles the biological relative abundance of different lipid types, with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids, depending on organism and organelle~\cite{vanmeer08}. Abbreviations and full names of all lipids present in the Databank are listed in Table~\ref{tab:abbreviations}. Force fields used in simulations cover all the essential parameter sets commonly used in lipid simulations, see Fig.~\ref{fig:overlay}C and Table~\ref{tab:ForceFields}, including also united atom and polarizable force fields. Therefore, the averages calculated over the Databank can be considered as mean predictions from available lipid models (average over force field parameters) for an average cell membrane (average over lipid compositions).

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay3.pdf}
    \caption{A) Schematic presentation of the overlay structure used in the NMRlipids Databank. A more detailed structure of the \textit{Databank layer} is shown in Fig.~\ref{DatabankStructure} in the SI.
    B) Distribution of lipids present in the trajectories of the Databank. `Others' lists lipids occurring in five or fewer simulations. 
    C) Distribution of force fields in the simulations in the Databank. References for each force field are given in Table~\ref{tab:ForceFields}.
    D) Flowchart for performing an analysis of properties through all MD simulations in the NMRlipids Databank using the API.
    E) Flowchart for accessing results calculated from the NMRlipids Databank and stored to the \textit{Application layer}.
    F) Currently available single lipid component and binary mixtures in the NMRlipids Databank. Colorbar shows the number of available simulations with the darkest green indicating three or more.  
    }
    \label{fig:overlay}
\end{figure}

\subsection{NMRlipids Databank-GUI: graphical access to the MD simulation data}
NMRlipids Databank-GUI, available at~\href{https://www.databank.nmrlipids.fi}{databank.nmrlipids.fi}, provides easy access to the NMRlipids Databank content through a graphical user interface (GUI). Simulations can be searched based on their molecular composition, force field, temperature, membrane properties, and quality; the search results are ranked based on the simulation quality as evaluated against experimental data. Membranes can be visualized, and properties between different simulations and experiments compared. The NMRlipids Databank-GUI enables rapid surveying of what simulation data is available, selection of the best available simulations for specific systems based on ranking lists, and comparisons of basic properties between different types of membranes. Notably, the GUI enables these operations to be performed by scientists with a wide range of backgrounds---including those who do not necessarily have programming expertise or other means to access MD simulation data.

\subsection{NMRlipids Databank-API: programmatic access to the MD simulation data}\label{section:access}
The NMRlipids Databank-API, available at \href{https://github.com/NMRlipids/Databank}{github.com/NMRlipids/Databank}, provides programmatic access to all simulation data in the NMRlipids Databank. This enables wide range of novel data-driven applications---from construction of machine learning models that predict membrane properties, to automatic analysis of virtually any property across all simulations in the Databank. The flowchart in Fig.~\ref{fig:overlay}D illustrates the practical implementation of such an analysis. First, raw data of each simulation can be accessed by iterating through the README.yaml files in the {\it Databank layer} that contain links pointing to the locations of simulation trajectories and other relevant files. Simulations can then be automatically analyzed with the help of the README.yaml and the mapping files that associate the specific naming conventions in each simulation with the universal molecule and atom names used by the Databank. Finally, the analysis results are stored using the same structure as in the {\it Databank layer}. Examples of codes that utilize the NMRlipids Databank-API, as well as a template for a new analysis, can be found from locations listed in Table~\ref{tab:codes}.

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{Figures/quality5.pdf}
    \caption{ A) Area per lipid of POPC and POPE lipid bilayers predicted by different force fields at 310\,K in simulations that are available in the NMRlipids Databank. The data points from the best-performing simulations, based on rankings in panels B and C, are surrounded by black circles. 
    B) Best POPC simulations ranked based on the \textit{sn}-1 acyl chain order parameter quality ($P^\mathrm{sn1}$). Also \textit{sn}-2 acyl chain ($P^\mathrm{sn2}$), headgroup ($P^\mathrm{hg}$) and total ($P^\mathrm{total}$) order parameter qualities, form factor quality ($FF_q$), and relative equilibration time for conformations ($\tau_\mathrm{rel}$) are shown.
    C) Best POPE simulations ranked based on the \textit{sn}-1 acyl chain order parameter quality. 
    D)--F) Direct comparison against experimental (NMR order parameters and X-ray scattering ) data exemplified for a simulation with the best overall order parameter quality (D), the best quality for POPE lipid (E), and the headgroup quality for POPE (F).
    G) Scatter plots and Pearson correlation coefficients, $r$, for the membrane area per lipid, thickness, first minimum of X-ray scattering form factor and average order parameter of the \textit{sn}-1 acyl chain extracted from the NMRlipids Databank. All correlation coefficients have p-value below 0.001. For more correlations see Fig.~\ref{fig:QualityCorrelationsSI}.
    }
    \label{fig:quality}
\end{figure}

\subsubsection{Detecting rare phenomena using NMRlipids Databank: Cholesterol flip-flops}
Lipid flip-flops from one bilayer leaflet to another play an important role in lipid trafficking and regulating membrane properties~\cite{vanmeer08}. Phospholipid flip-flop events are rare when not facilitated by proteins, occurring spontaneously on the timescale of hours or days, while cholesterol, diacylglycerol, and ceramide flip-flop much more often. Still, the reported timescales range from minutes to sub-millisecods~\cite{vanmeer08,steck12,parisio16,gu19}. These timescales were previously accessible only by coarse-grained simulations or free energy calculations~\cite{parisio16}, and atomistic simulations reporting cholesterol flip-flop events have been published only recently~\cite{gu19,javanainen19,baral20}. The atomistic studies report an increase in cholesterol flip-flop rates with increasing acyl chain unsaturation level and decreasing cholesterol concentration~\cite{gu19,javanainen19}, but the amount of data in these individual studies was not sufficient to systematically assess correlations between cholesterol flip-flop rates and membrane properties. Here, we demonstrate that the NMRlipids Databank-API makes analyses of such rare phenomena accessible for all by enabling access to a large amount of MD simulation data as described in section~\ref{section:access}. This is particularly useful for scientists in various fields of science and industry who lack access to the computational resources or the expertise to produce the large amounts of MD simulation data required for such analyses.

Using the general workflow depicted in Fig.~\ref{fig:overlay}D, we first calculated the flip-flop rates from all the simulations available in the NMRlipids Databank. Flip-flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-\textit{sn}-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-\textit{sn}-glycerol).
The observed cholesterol flip-flop rates, ranging between 0.001--1.6\,\textmu{}s$^{-1}$ with the mean of 0.16\,\textmu{}s$^{-1}$ and median of 0.07\,\textmu{}s$^{-1}$, are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}.
The flip-flop rate of DCHOL, 0.2\,\textmu{}s$^{-1}$, was close to the average value of cholesterol, while
the average rates for diacylglycerols DOG (0.4\,\textmu{}s$^{-1}$) and SDG (0.5\,\textmu{}s$^{-1}$) were higher than for cholesterol.
Flip-flops were not observed for other lipids, giving the upper limits for
PC-lipid flip-flop rate as 9$\times$10$^{-6}$\,\textmu{}s$^{-1}$ and for
ceramide (\textit{N}-palmitoyl-\textsc{D}-\textit{erythro}-sphingosine) as 0.002\,\textmu{}s$^{-1}$.
Thus, the available data in the NMRlipids Databank suggest that the lipid flip-flop rate decreases in the order: diacylglycerols $>$ cholesterol $>$ other lipids including ceramides.
However, the amount of data for diacylgycerols (8 simulations with the Lipid17 force field) and ceramide (3 simulations with CHARMM36) is less than that for cholesterol (83 simulations); thus we cannot fully exclude the effect of force field or composition on this comparison.

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=88mm]{Figures/CholFlipFlops.pdf}
    \caption{\textbf{A} Illustration of cholesterol flip-flop.  
      \textbf{B--D} Cholesterol flip-flops analyzed from the Databank as a function of membrane thickness, area per lipid, and acyl chain order. Values from simulations with non-zero flip-flop rates are shown with blue dots. Histogrammed values are shown with black dots. For the mean value in each bin, average weighted with the simulation lengths was used, and error bars show the standard error of the mean.
    }
    \label{fig:flip-flops}
\end{figure}

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=\linewidth]{Figures/permeation3.pdf}
    \caption{\textbf{A} Water diffusion, $D_\perp$, and permeability, $P$, through membranes, and lateral diffusion along the membrane, $D_\parallel$, illustrated in a multilamellar stack of lipid bilayers. 
    \textbf{B--E} Water permeation through membranes analyzed from the Databank as a function of temperature, thickness, area per lipid, and acyl chain order. Inset in B) shows the Arrhenius plot of permeation ($\ln(P)$ vs. $1/T$) that gives 14\,$\pm$\,3\,$k_\mathrm{B}T$ for the average activation energy for water permeation through lipid bilayer. 
    \textbf{F} Lateral diffusion of water as a function of hydration level. Experimental points for DMPC bilayers at 313\,K at different hydration levels are shown~\cite{rudakova04}.
    \textbf{G--H} Diffusion anistoropy of water as a function of thickness and area per lipid. 
    Non-zero permeation and diffusion values from simulations are shown with blue dots. Histogrammed values are shown with black dots. For the mean value in each bin, average weighted with the simulation lengths was used, and error bars show the standard error of the mean. Only bins with more than one microsecond of data in total were used for water permeation.}
    \label{fig:permeability}
\end{figure}

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids Databank using the general workflow depicted in Fig.~\ref{fig:overlay}D. The resulting non-zero values range between 0.3--322\,\textmu{}m/s with the mean of 14\,\textmu{}m/s and median of 8\,\textmu{}m/s. These values agree with the previously reported simulation results~\cite{venable19,camilo2022}, but are on average slightly larger than experimental values reported for PC lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Using the workflow depicted in Fig.~\ref{fig:overlay}E, we then plotted the observed permeabilities and their histogrammed values in Figs.~\ref{fig:permeability}B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 14\,$\pm$\,3\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability}B. On the other hand, the water permeability on average decreases when membranes become more packed, that is, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability}C--E). Permeation of water through bilayers depends on membrane properties also according to previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids Databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters.
%, yet the linear correlation would be a good approximation for thicknesses $\gtrsim$\,3.9\,nm and area per lipids $\lesssim$\,69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability}C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on hydration level or the fraction of charged lipids, cholesterol, or POPE in the membrane were not observed (Fig.~\ref{fig:permeationSI}).
%
%%Several models on how water dynamics depends on membrane properties have been proposed~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids Databank using the general workflow depicted in Fig.~\ref{fig:overlay}D. The parallel diffusion coefficient of water, $D_\parallel$, decreases with reduced hydration and increases with the temperature, but dependencies on the membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI} when the workflow depicted in Fig.~\ref{fig:overlay}E was used. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability}F, but increase to approximately two times higher than the experimental value for bulk water diffusion value (3.1$\times$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels.
This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\perp/D_\parallel$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\perp$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than the lateral diffusion coefficients of water (Figs.~\ref{fig:permeability}G--H), which is at the upper limit of anisotropy estimated from experimental data~\cite{nitsche19}. A significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\perp/D_\parallel$ deviates further from unity with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability}G--H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability}C--D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI}A,~C).

@@

The open programmatic access to unprecedented amounts of MD simulation data through the NMRlipids Databank-API enables a wide range of novel data-driven analyses and machine learning applications. The analyses of cholesterol flip-flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}) demonstrate how a large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enable analyses of rare phenomena that are beyond the current possibilities for a single research group. Such analyses also pave the way for applications of MD simulations in new fields, as demonstrated here by analysing an essential parameter in pharmacokinetic modeling and MRI imaging:\cite{nitsche19,topgaard20} the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}). Furthermore, the NMRlipids Databank-API delivers access to a training set data that paves the way for diverse machine learning applications to 
%For example, a machine learning model predicting electron density profiles from form factors would be highly useful in interpretation of scattering experiments.
%In addition, the data contained in the NMRlipids databank will enable training of more elaborated models to
predict membrane properties. Such applications are analogous to AlphaFold \cite{jumper21} and other tools that predict protein structures from their sequence using artificial intelligence. These possibilities are particularly valuable for scientists who do not typically have access to large scale MD simulation data. Expected applications of the NMRlipids Databank in a wide range of fields are listed in Table~\ref{tab:applications}, yet the scope of such applications
%in biology, biotechnology, material science, biomolecular imaging and beyond
is expected to further widen with increasing amount of data in the Databank.

@@

\subsection{Calculation of C--H bond order parameters}
The C--H bond order parameters were calculated directly from the carbon and hydrogen positions using the definition
%The order parameters are defined as
\begin{equation}\label{OP}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where angular brackets denote the ensemble average, \textit{i.e.}, average over all sampled configurations of all lipids in a simulation, and $\theta$ is the angle between the C--H bond and the membrane normal. As in previous NMRlipids publications, the order parameters were first calculated separately for each lipid and the standard error of the mean over different lipids was used as the error estimate~\cite{botan15}. However, order parameters for simulations with $\tau_{\rm rel} > 1$ may be influenced by the starting structure and thereby their error bars may be underestimated. The script that calculates C--H bond order parameters from all simulations in the NMRlipids Databank is available at \texttt{/Scripts/AnalyzeDatabank/calcOrderParameters.py} in the NMRlipids Databank repository. The resulting order parameters are stored for all simulations in files named %\newline 
[\texttt{lipid\_name}]\texttt{OrderParameters.json} at folders in \texttt{/Data/Simulations} in the NMRlipids Databank repository.

>>>>

NMRlipids Databank is a community-driven catalogue containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. It has been designed to improve the Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} of MD simulation data, most importantly the output trajectories and necessary information to their reuse. The NMRlipids Databank is constructed using the NMRlipids project protocol, in which all the content is openly accessible throughout the project~\cite{botan15}. Currently, the NMRlipids Databank contains 726 simulation trajectories with the total length of approximately 0.4\,ms. Single-component lipid membranes and binary mixtures, depicted in Fig.~\ref{fig:overlay}F, are currently most abundant in the NMRlipids Databank, yet mixtures with up to five lipid types are available. The distribution of lipids among the available simulations, shown in Fig.~\ref{fig:overlay}B, roughly resembles the biological relative abundance of different lipid types, with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids, depending on organism and organelle~\cite{vanmeer08}. Abbreviations and full names of all lipids present in the Databank are listed in Table~\ref{tab:abbreviations}. Force fields used in simulations cover all the essential parameter sets commonly used in lipid simulations, see Fig.~\ref{fig:overlay}C and Table~\ref{tab:ForceFields}, including also united atom and polarizable force fields. Therefore, the averages calculated over the Databank can be considered as mean predictions from available lipid models (average over force field parameters) for an average cell membrane (average over lipid compositions).

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay3.pdf}
    \caption{Overview of the NMRlipids Databank. 
    \textbf{A} Schematic presentation of the overlay structure used in the NMRlipids Databank. A more detailed structure of the \textit{Databank layer} is shown in Fig.~\ref{DatabankStructure} in the SI.
    \textbf{B} Distribution of lipids present in the trajectories of the Databank. `Others' lists lipids occurring in five or fewer simulations. 
    \textbf{C} Distribution of force fields in the simulations in the Databank. References for each force field are given in Table~\ref{tab:ForceFields}.
    \textbf{D} Flowchart for performing an analysis of properties through all MD simulations in the NMRlipids Databank using the API.
    \textbf{E} Flowchart for accessing results calculated from the NMRlipids Databank and stored to the \textit{Application layer}.
    \textbf{F} Currently available single lipid component and binary mixtures in the NMRlipids Databank. Colorbar shows the number of available simulations with the darkest green indicating three or more.  
    }
    \label{fig:overlay}
\end{figure}

\subsection{NMRlipids Databank-GUI: graphical access to the MD simulation data}
NMRlipids Databank-GUI, available at~\href{https://www.databank.nmrlipids.fi}{databank.nmrlipids.fi}, provides easy access to the NMRlipids Databank content through a graphical user interface (GUI). Simulations can be searched based on their molecular composition, force field, temperature, membrane properties, and quality; the search results are ranked based on the simulation quality as evaluated against experimental data when available. Membranes can be visualized, and properties between different simulations and experiments compared. The NMRlipids Databank-GUI enables rapid surveying of what simulation data is available, selection of the best available simulations for specific systems based on ranking lists, and comparisons of basic properties between different types of membranes. Notably, the GUI enables these operations to be performed by scientists with a wide range of backgrounds---including those who do not necessarily have programming expertise or other means to access MD simulation data.

\subsection{NMRlipids Databank-API: programmatic access to the MD simulation data}\label{section:access}
The NMRlipids Databank-API, available at \href{https://github.com/NMRlipids/Databank}{github.com/NMRlipids/Databank}, provides programmatic access to all simulation data in the NMRlipids Databank. This enables wide range of novel data-driven applications---from construction of machine learning models that predict membrane properties, to automatic analysis of virtually any property across all simulations in the Databank. The flowchart in Fig.~\ref{fig:overlay}D illustrates the practical implementation of such an analysis. After cloning the databank repository to a local computer, raw data of each simulation can be accessed by iterating through the README.yaml files in the {\it Databank layer} that contain links pointing to the locations of simulation trajectories and other relevant files. Simulations can then be automatically analyzed with the help of the README.yaml and the mapping files that associate the specific naming conventions in each simulation with the universal molecule and atom names used by the Databank. Finally, the analysis results are stored using the same structure as in the {\it Databank layer}. Practical examples of codes that utilize the NMRlipids Databank-API, as well as a template for a new analysis, can be found from locations listed in Table~\ref{tab:codes}.

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{Figures/quality5.pdf}
    \caption{Examples of data obtained using NMRlipids Databank. 
    \textbf{A} Area per lipid of POPC and POPE lipid bilayers predicted by different force fields at 310\,K in simulations that are available in the NMRlipids Databank. The data points from the best-performing simulations, based on rankings in panels B and C, are surrounded by black circles. 
    \textbf{B} Best POPC simulations ranked based on the \textit{sn}-1 acyl chain order parameter quality ($P^\mathrm{sn1}$). Also \textit{sn}-2 acyl chain ($P^\mathrm{sn2}$), headgroup ($P^\mathrm{hg}$) and total ($P^\mathrm{total}$) order parameter qualities, form factor quality ($FF_q$), and relative equilibration time for conformations ($\tau_\mathrm{rel}$) are shown. Note that the best possible order parameter quality is one, while the best possible form factor quality is zero.
    \textbf{C} Best POPE simulations ranked based on the \textit{sn}-1 acyl chain order parameter quality. 
    \textbf{D--F} Direct comparison against experimental (NMR order parameters and X-ray scattering ) data exemplified for a simulation with the best overall order parameter quality (D), the best quality for POPE lipid (E), and the headgroup quality for POPE (F).
    \textbf{G} Scatter plots and Pearson correlation coefficients, $r$, for the membrane area per lipid, thickness, first minimum of X-ray scattering form factor and average order parameter of the \textit{sn}-1 acyl chain extracted from the NMRlipids Databank. All correlation coefficients have p-value below 0.001. For more correlations see Fig.~\ref{fig:QualityCorrelationsSI}.
    }
    \label{fig:quality}
\end{figure}

\subsubsection{Detecting rare phenomena using NMRlipids Databank: Cholesterol flip-flops}
Lipid flip-flops from one bilayer leaflet to another play an important role in lipid trafficking and regulating membrane properties~\cite{vanmeer08}. Phospholipid flip-flop events are rare when not facilitated by proteins, occurring spontaneously on the timescale of hours or days, while cholesterol, diacylglycerol, and ceramide flip-flop much more often. Still, the reported timescales range from minutes to sub-millisecods~\cite{vanmeer08,steck12,parisio16,gu19}. These timescales were previously accessible only by coarse-grained simulations or free energy calculations~\cite{parisio16}, and atomistic simulations reporting cholesterol flip-flop events have been published only recently~\cite{gu19,javanainen19,baral20}. The atomistic studies report an increase in cholesterol flip-flop rates with increasing acyl chain unsaturation level and decreasing cholesterol concentration~\cite{gu19,javanainen19}, but the amount of data in these individual studies was not sufficient to systematically assess correlations between cholesterol flip-flop rates and membrane properties. Here, we demonstrate that the NMRlipids Databank-API makes analyses of such rare phenomena accessible for all by enabling access to a large amount of MD simulation data as illustrated in Fig.~\ref{fig:overlay}. This is particularly useful for scientists in various fields of science and industry who lack access to the computational resources or the expertise to produce the large amounts of MD simulation data required for such analyses.

Using the general workflow depicted in Fig.~\ref{fig:overlay}D, we first calculated the flip-flop rates from all the simulations available in the NMRlipids Databank. Flip-flops were observed for cholesterol, DCHOL (18,19-di-nor-cholesterol), DOG (1,2-dioleoyl-\textit{sn}-glycerol), and SDG (1-stearoyl-2-docosahexaenoyl-\textit{sn}-glycerol). The observed cholesterol flip-flop rates, ranging between 0.001--1.6\,\textmu{}s$^{-1}$ with the mean of 0.16\,\textmu{}s$^{-1}$ and median of 0.07\,\textmu{}s$^{-1}$, are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. The flip-flop rate of DCHOL, 0.2\,\textmu{}s$^{-1}$, was close to the average value of cholesterol, while the average rates for diacylglycerols DOG (0.4\,\textmu{}s$^{-1}$) and SDG (0.5\,\textmu{}s$^{-1}$) were higher than for cholesterol. Flip-flops were not observed for other lipids, giving the upper limits for PC-lipid flip-flop rate as 9$\times$10$^{-6}$\,\textmu{}s$^{-1}$ and for ceramide (\textit{N}-palmitoyl-\textsc{D}-\textit{erythro}-sphingosine) as 0.002\,\textmu{}s$^{-1}$. Thus, the available data in the NMRlipids Databank suggest that the lipid flip-flop rate decreases in the order: diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, the amount of data for diacylgycerols (8 simulations with the Lipid17 force field) and ceramide (3 simulations with CHARMM36) is less than that for cholesterol (83 simulations); thus we cannot fully exclude the effect of force field or composition on this comparison.

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=88mm]{Figures/CholFlipFlops.pdf}
    \caption{Quantification of cholesterol flip-flop events in NMRlipids Databank simulations. 
    \textbf{A} Illustration of cholesterol flip-flop.  
      \textbf{B--D} Cholesterol flip-flops analyzed from the Databank as a function of membrane thickness, area per lipid, and acyl chain order. Values from simulations with non-zero flip-flop rates are shown with blue dots. Histogrammed values are shown with black dots. For the mean value in each bin, average weighted with the simulation lengths was used, and error bars show the standard error of the mean.
    }
    \label{fig:flip-flops}
\end{figure}

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=\linewidth]{Figures/permeation3.pdf}
    \caption{Quantification of water diffusion in NMRlipids Databank simulations.
    \textbf{A} Water diffusion, $D_\perp$, and permeability, $P$, through membranes, and lateral diffusion along the membrane, $D_\parallel$, illustrated in a multilamellar stack of lipid bilayers. 
    \textbf{B--E} Water permeation through membranes analyzed from the Databank as a function of temperature, thickness, area per lipid, and acyl chain order. Inset in B) shows the Arrhenius plot of permeation ($\ln(P)$ vs. $1/T$) that gives 14\,$\pm$\,3\,$k_\mathrm{B}T$ for the average activation energy for water permeation through lipid bilayer. 
    \textbf{F} Lateral diffusion of water as a function of hydration level. Experimental points for DMPC bilayers at 313\,K at different hydration levels are shown~\cite{rudakova04}.
    \textbf{G--H} Diffusion anistoropy of water as a function of thickness and area per lipid. 
    Non-zero permeation and diffusion values from simulations are shown with blue dots. Histogrammed values are shown with black dots. For the mean value in each bin, average weighted with the simulation lengths was used, and error bars show the standard error of the mean. Only bins with more than one microsecond of data in total were used for water permeation.}
    \label{fig:permeability}
\end{figure}

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids Databank using the general workflow depicted in Fig.~\ref{fig:overlay}D. The resulting non-zero values range between 0.3--322\,\textmu{}m/s with the mean of 14\,\textmu{}m/s and median of 8\,\textmu{}m/s. These values agree with the previously reported simulation results~\cite{venable19,camilo2022}, but are on average larger than experimental values reported for PC lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Using the workflow depicted in Fig.~\ref{fig:overlay}E, we then plotted the observed permeabilities and their histogrammed values in Figs.~\ref{fig:permeability}B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 14\,$\pm$\,3\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability}B. On the other hand, the water permeability on average decreases when membranes become more packed, that is, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability}C--E). Permeation of water through bilayers depends on membrane properties also according to previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids Databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters.
%, yet the linear correlation would be a good approximation for thicknesses $\gtrsim$\,3.9\,nm and area per lipids $\lesssim$\,69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability}C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on hydration level or the fraction of charged lipids, cholesterol, or POPE in the membrane were not observed (Fig.~\ref{fig:permeationSI}).
%
%%Several models on how water dynamics depends on membrane properties have been proposed~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To examine how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we analyzed the water diffusion parallel to the membrane surface from all simulations in the NMRlipids Databank using the general workflows depicted in Figs.~\ref{fig:overlay}D and E. The parallel diffusion coefficient of water, $D_\parallel$, decreases with reduced hydration and increases with the temperature, but dependencies on the membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI}. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability}F, but increase to approximately 50\% higher than the experimental value for bulk water diffusion value (3.1$\times$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels.
This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\perp/D_\parallel$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\perp$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than the lateral diffusion coefficients of water (Figs.~\ref{fig:permeability}G--H), which is at the upper limit of anisotropy estimated from experimental data~\cite{nitsche19}. A significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\perp/D_\parallel$ deviates further from unity with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability}G--H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability}C--D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI}A,~C).

@@

The open programmatic access to unprecedented amounts of MD simulation data through the NMRlipids Databank-API enables a wide range of novel data-driven analyses and machine learning applications. The analyses of cholesterol flip-flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}) demonstrate how a large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enable analyses of rare phenomena that are beyond the current possibilities for a single research group. Such analyses also pave the way for applications of MD simulations in new fields, as demonstrated here by analysing an essential parameter in pharmacokinetic modeling and MRI imaging:\cite{nitsche19,topgaard20} the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}). Furthermore, the NMRlipids Databank-API delivers access to a training set data that paves the way for diverse machine learning applications to predict membrane properties. Such applications could be analogous to AlphaFold \cite{jumper21} and other tools that predict protein structures from their sequence using artificial intelligence. These possibilities are particularly valuable for scientists who do not typically have access to large scale MD simulation data. Expected applications of the NMRlipids Databank in a wide range of fields are listed in Table~\ref{tab:applications}, yet the scope of such applications
%in biology, biotechnology, material science, biomolecular imaging and beyond
is expected to further widen with increasing amount of data in the Databank.

@@

\subsection{Calculation of C--H bond order parameters}
The C--H bond order parameters were calculated directly from the carbon and hydrogen positions using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where angular brackets denote the ensemble average, \textit{i.e.}, average over all sampled configurations of all lipids in a simulation, and $\theta$ is the angle between the C--H bond and the membrane normal. As in previous NMRlipids publications, the order parameters were first calculated separately for each lipid and the standard error of the mean over different lipids was used as the error estimate~\cite{botan15}. However, order parameters for simulations with $\tau_{\rm rel} > 1$ may be influenced by the starting structure and thereby their error bars may be underestimated. The script that calculates C--H bond order parameters from all simulations in the NMRlipids Databank is available at \texttt{/Scripts/AnalyzeDatabank/calcOrderParameters.py} in the NMRlipids Databank repository. The resulting order parameters are stored for all simulations in files named %\newline 
[\texttt{lipid\_name}]\texttt{OrderParameters.json} at folders in \texttt{/Data/Simulations} in the NMRlipids Databank repository.
